Spatiotemporal Multi-Resolution Approximation of the Amari Type Neural Field Model

### 2012-01-31 Multiresolution_framework/ltx/manuscript.tex
edits to lower bound appendix

\subsection{Lower bound on likelihood function}\label{ap:MStep}
A derivation of the lower bound on the likelihood function that is based on \citet{Bishop2006} is provided in this Appendix. Let the entire sequence of observations is denoted by $\mathbf Y$, and the sequence of hidden variables by $\mathbf X$ and the parameter set by $\boldsymbol\Theta$. We want to maximise the likelihood function which is difficult to maximise directly
\begin{equation}
 p(\mathbf Y;\boldsymbol\Theta)=\int_{\mathbf X}p(\mathbf X,\mathbf Y;\boldsymbol\Theta)d\mathbf X.
\end{equation}   
Maximising the likelihood in terms of the parameter set is equivalent to the log-likelihood maximisation, as both the likelihood and log-likelihood functions share the same maximiser due to the monotonicity of the logarithm function. For any choice of distribution defined over the latent variables, $q(\mathbf X)$, the following decomposition holds
\begin{equation}\label{eq:EM-loglikelihooddecomposition}
 \ln~p(\mathbf Y;\boldsymbol\Theta)=\mathcal{L}(q,\boldsymbol\Theta)+\mathrm{KL}(q\parallel p),
\end{equation}
where
\begin{align}
 \mathcal{L}(q,\boldsymbol\Theta)&=\int_{\mathbf X} q(\mathbf X)\ln\left\lbrace \frac{p(\mathbf X,\mathbf Y;\boldsymbol\Theta)}{q(\mathbf X)} \right\rbrace d\mathbf X \label{eq:EM-decomposition1}\\
\mathrm{KL}(q\parallel p)&=-\int_{\mathbf X}q(\mathbf X)\ln\left\lbrace \frac{p(\mathbf X \mid \mathbf Y;\boldsymbol\Theta)}{q(\mathbf X)} \right\rbrace d\mathbf X. \label{eq:EM-decomposition2}
\end{align}
Equation \eqref{eq:EM-loglikelihooddecomposition} can be verified by applying the probability product rule to the joint distribution $p(\mathbf X,\mathbf Y;\boldsymbol\Theta)$ in \eqref{eq:EM-decomposition1} giving
\begin{align}
	\mathcal{L}(q,\boldsymbol\Theta)+\mathrm{KL}(q\parallel p)&= \int_{\mathbf X} q(\mathbf X)\ln\left\lbrace \frac{p(\mathbf X \mid \mathbf Y;\boldsymbol\Theta)p(\mathbf Y;\boldsymbol\Theta)}{q(\mathbf X)} \right\rbrace d\mathbf X \nonumber\\
	&- \int_{\mathbf X}q(\mathbf X)\ln\left\lbrace \frac{p(\mathbf X \mid \mathbf Y;\boldsymbol\Theta)}{q(\mathbf X)} \right\rbrace d\mathbf X\nonumber\\
	&= \int_{\mathbf X}q(\mathbf X)\ln p(\mathbf Y; \boldsymbol\theta)d\mathbf X.
\end{align}
Noting  $q(\mathbf X)$ is a normalised distribution that integrates to one gives the log-likelihood function. Note \eqref{eq:EM-decomposition2} is the Kullback-Liebler distance between distributions $p(\mathbf X \mid \mathbf Y;\boldsymbol\Theta)$ and $q(\mathbf X)$, therefore,
\begin{equation}
	\mathrm{KL}(q\parallel p)\ge 0.
\end{equation}   
The equality holds if and only if $q(\mathbf X)=p(\mathbf X \mid \mathbf Y;\boldsymbol\Theta)$. It follows from equation \eqref{eq:EM-loglikelihooddecomposition}, that the quantity $\mathcal{L}(q,\boldsymbol\Theta)$ is a lower bound on the log-likelihood function, i.e.
\begin{equation}
 \mathcal{L}(q,\boldsymbol\Theta)\le \ln~p(\mathbf Y\mid\boldsymbol\Theta)
	\end{equation}

The EM algorithm alternates between maximising the lower bound with respect to the distribution $q(\mathbf X)$ and the parameter set $\boldsymbol\Theta$, respectively, while holding the other fixed. Suppose that the current estimate of the parameter set is $\boldsymbol\Theta'$, in the E-step the lower bound is maximised with respect to the distribution $q(\mathbf X)$ while $\boldsymbol\Theta'$ is fixed. Since the value of $p(\mathbf Y;\boldsymbol\Theta')$ is independent of $q(\mathbf X)$ the maximum of the lower bound occurs when the KL distance disappears or in other words when $q(\mathbf X)$ is exactly the conditional distribution $p(\mathbf X \mid \mathbf Y;\boldsymbol\Theta)$. At this point, the lower bound becomes equal to the log-likelihood function. Note in the state-space model the E-step is equivalent to the smoothing problem. The maximum in the M-step is obtained by maximising the lower bound with respect to the parameter set while the distribution $q(\mathbf X)$ is held fixed. Since the distribution $q(\mathbf X)$ is computed using the old parameter estimates and kept fixed during the M-step, it is not equal to $p(\mathbf X \mid \mathbf Y;\boldsymbol\Theta)$, resulting in a non-zero KL distance. Therefore, the growth of the log-likelihood function is greater than the growth of the lower bound. Substituting $p(\mathbf X \mid \mathbf Y;\boldsymbol\Theta')$ for $q(\mathbf X)$ in \eqref{eq:EM-decomposition1} gives
\begin{align}
 \mathcal{L}(q,\boldsymbol\Theta)&=\int_{\mathbf X} p(\mathbf X\mid\mathbf Y;\boldsymbol\Theta')\ln p(\mathbf X,\mathbf Y;\boldsymbol\Theta)~d\mathbf X-\int_{\mathbf X} p(\mathbf X\mid\mathbf Y;\boldsymbol\Theta')\ln p(\mathbf X\mid\mathbf Y;\boldsymbol\Theta')~d\mathbf X \nonumber \\
&= \mathbf E_{\boldsymbol\Theta'}\left[ \ln p(\mathbf X,\mathbf Y;\boldsymbol\Theta)\right] +\mathrm{const.}\nonumber\\
&=\mathcal{Q}(\boldsymbol\Theta,\boldsymbol\Theta')+\mathrm{const.}~,\label{eq:EM-ComDaLogLike}
\end{align}
where $ \mathbf E_{\boldsymbol \Theta'}\left[ .\right] $ denotes expectation taken with respect to the marginal distribution of the states conditioned on the observed field and the current estimate of the parameter set, i.e.  $p(\mathbf X\mid\mathbf Y;\boldsymbol \Theta'$). The constant term is the negative entropy of $q(\mathbf X)$ and is independent of $\boldsymbol\Theta$.   It becomes clear from \eqref{eq:EM-ComDaLogLike} that, the M-step essentially maximises the expectation of the complete data log-likelihood. 
\section{Calculation of $\boldsymbol\Xi_{0}$ and $\boldsymbol\Xi_{1}$}
The expressions for $\boldsymbol\Xi_{0}$ and $\boldsymbol\Xi_{1}$ in terms of smoother outputs can be found in \cite{Shumway2000}, here we present the derivation for completeness.  The expression for  $\boldsymbol\Xi_{0}$ as defined in equation~\eqref{eq:Xi0} of the main text is given by
   \begin{equation}\label{eq:appXi0}
	 \boldsymbol\Xi_0=\mathbf E_{\theta'}\left[\sum_{t=0}^{T-1}\mathbf x_t\mathbf x_{t+1}^\top\right].
	\end{equation}                                                                                                
	We start by expanding the equation \eqref{eq:appXi0}, giving

>>>>

\subsection{Lower bound on likelihood function}\label{ap:MStep}
A derivation of the lower bound on the likelihood function that is based on \citet{Bishop2006} is provided in this Appendix. Let the entire sequence of observations is denoted by $\mathbf Y$, and the sequence of hidden variables by $\mathbf X$ and the parameter set by $\boldsymbol\theta$. We want to maximise the likelihood function which is difficult to maximise directly
\begin{equation}
 p(\mathbf Y;\boldsymbol\theta)=\int_{\mathbf X}p(\mathbf X,\mathbf Y;\boldsymbol\theta)d\mathbf X.
\end{equation}   
Maximising the likelihood in terms of the parameter set is equivalent to the log-likelihood maximisation, as both the likelihood and log-likelihood functions share the same maximiser due to the monotonicity of the logarithm function. For any choice of distribution defined over the latent variables, $q(\mathbf X)$, the following decomposition holds
\begin{equation}\label{eq:EM-loglikelihooddecomposition}
 \ln~p(\mathbf Y;\boldsymbol\theta)=\mathcal{L}(q,\boldsymbol\theta)+\mathrm{KL}(q\parallel p),
\end{equation}
where
\begin{align}
 \mathcal{L}(q,\boldsymbol\theta)&=\int_{\mathbf X} q(\mathbf X)\ln\left\lbrace \frac{p(\mathbf X,\mathbf Y;\boldsymbol\theta)}{q(\mathbf X)} \right\rbrace d\mathbf X \label{eq:EM-decomposition1}\\
\mathrm{KL}(q\parallel p)&=-\int_{\mathbf X}q(\mathbf X)\ln\left\lbrace \frac{p(\mathbf X \mid \mathbf Y;\boldsymbol\theta)}{q(\mathbf X)} \right\rbrace d\mathbf X. \label{eq:EM-decomposition2}
\end{align}
Equation \eqref{eq:EM-loglikelihooddecomposition} can be verified by applying the probability product rule to the joint distribution $p(\mathbf X,\mathbf Y;\boldsymbol\theta)$ in \eqref{eq:EM-decomposition1} giving
\begin{align}
	\mathcal{L}(q,\boldsymbol\theta)+\mathrm{KL}(q\parallel p)&= \int_{\mathbf X} q(\mathbf X)\ln\left\lbrace \frac{p(\mathbf X \mid \mathbf Y;\boldsymbol\theta)p(\mathbf Y;\boldsymbol\theta)}{q(\mathbf X)} \right\rbrace d\mathbf X \nonumber\\
	&- \int_{\mathbf X}q(\mathbf X)\ln\left\lbrace \frac{p(\mathbf X \mid \mathbf Y;\boldsymbol\theta)}{q(\mathbf X)} \right\rbrace d\mathbf X\nonumber\\
	&= \int_{\mathbf X}q(\mathbf X)\ln p(\mathbf Y; \boldsymbol\theta)d\mathbf X.
\end{align}
Noting  $q(\mathbf X)$ is a normalised distribution that integrates to one gives the log-likelihood function. Note \eqref{eq:EM-decomposition2} is the Kullback-Liebler distance between distributions $p(\mathbf X \mid \mathbf Y;\boldsymbol\theta)$ and $q(\mathbf X)$, therefore,
\begin{equation}
	\mathrm{KL}(q\parallel p)\ge 0.
\end{equation}   
The equality holds if and only if $q(\mathbf X)=p(\mathbf X \mid \mathbf Y;\boldsymbol\theta)$. It follows from equation \eqref{eq:EM-loglikelihooddecomposition}, that the quantity $\mathcal{L}(q,\boldsymbol\theta)$ is a lower bound on the log-likelihood function, i.e.
\begin{equation}
 \mathcal{L}(q,\boldsymbol\theta)\le \ln~p(\mathbf Y\mid\boldsymbol\theta)
	\end{equation}

The EM algorithm alternates between maximising the lower bound with respect to the distribution $q(\mathbf X)$ and the parameter set $\boldsymbol\theta$, respectively, while holding the other fixed. Suppose that the current estimate of the parameter set is $\boldsymbol\theta'$, in the E-step the lower bound is maximised with respect to the distribution $q(\mathbf X)$ while $\boldsymbol\theta'$ is fixed. Since the value of $p(\mathbf Y;\boldsymbol\theta')$ is independent of $q(\mathbf X)$ the maximum of the lower bound occurs when the KL distance disappears or in other words when $q(\mathbf X)$ is exactly the conditional distribution $p(\mathbf X \mid \mathbf Y;\boldsymbol\theta)$. At this point, the lower bound becomes equal to the log-likelihood function. Note in the state-space model the E-step is equivalent to the smoothing problem. The maximum in the M-step is obtained by maximising the lower bound with respect to the parameter set while the distribution $q(\mathbf X)$ is held fixed. Since the distribution $q(\mathbf X)$ is computed using the old parameter estimates and kept fixed during the M-step, it is not equal to $p(\mathbf X \mid \mathbf Y;\boldsymbol\theta)$, resulting in a non-zero KL distance. Therefore, the growth of the log-likelihood function is greater than the growth of the lower bound. Substituting $p(\mathbf X \mid \mathbf Y;\boldsymbol\theta')$ for $q(\mathbf X)$ in \eqref{eq:EM-decomposition1} gives
\begin{align}
 \mathcal{L}(q,\boldsymbol\theta)&=\int_{\mathbf X} p(\mathbf X\mid\mathbf Y;\boldsymbol\theta')\ln p(\mathbf X,\mathbf Y;\boldsymbol\theta)~d\mathbf X-\int_{\mathbf X} p(\mathbf X\mid\mathbf Y;\boldsymbol\theta')\ln p(\mathbf X\mid\mathbf Y;\boldsymbol\theta')~d\mathbf X \nonumber \\
&= \mathbf E_{\boldsymbol\theta'}\left[ \ln p(\mathbf X,\mathbf Y;\boldsymbol\theta)\right] +\mathrm{const.}\nonumber\\
&=\mathcal{Q}(\boldsymbol\theta,\boldsymbol\theta')+\mathrm{const.}~,\label{eq:EM-ComDaLogLike}
\end{align}
where $ \mathbf E_{\boldsymbol \theta'}\left[ \cdot\right] $ denotes expectation taken with respect to the marginal distribution of the states conditioned on the observed field and the current estimate of the parameter set, i.e.  $p(\mathbf X\mid\mathbf Y;\boldsymbol \theta'$). The constant term is the negative entropy of $q(\mathbf X)$ and is independent of $\boldsymbol\theta$.   It becomes clear from \eqref{eq:EM-ComDaLogLike} that, the M-step essentially maximises the expectation of the complete data log-likelihood. 
\section{Calculation of $\boldsymbol\Xi_{0}$ and $\boldsymbol\Xi_{1}$}
The expressions for $\boldsymbol\Xi_{0}$ and $\boldsymbol\Xi_{1}$ in terms of smoother outputs can be found in \cite{Shumway2000}, here we present the derivation for completeness.  The expression for  $\boldsymbol\Xi_{0}$ as defined in equation~\eqref{eq:Xi0} of the main text is given by
   \begin{equation}\label{eq:appXi0}
	 \boldsymbol\Xi_0=\mathbf E_{\theta'}\left[\sum_{t=0}^{T-1}\mathbf x_t\mathbf x_{t+1}^\top\right].
	\end{equation}                                                                                                
	We start by expanding the equation \eqref{eq:appXi0}, giving
